Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-26 draft.tex
All sections are written, Images still not done

%% \section{Introduction} %for journal use above \firstsection{..} instead
\textbf{(Background)} Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
On the other hand, observations made by clinicians in the daily routine are translated into hypothesis for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical attributes which may influence a condition of interest.
%
The data variables necessary are gathered using structured interviews and clinical examinations.
%
Statistical methods like regression analysis aim to check the attribute list for plausibility.
%
\\\\
\textbf{(Problem)}
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Modern cohort studies often include medical image data.
%
These data needs to be segmented to analyze its features for correlations with non-image data.
%
Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques are more promising but challenging since the used methods as MRI and ultrasound are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.

@@

\textbf{(Goals)}

@@

\textbf{(Contributions)}

@@

Morphometric variables are derived to allow for statistical comparison of the tissue which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed Version of the \emph{IVA Loop}. Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types. The data is visualized in the range and domain space, which can then be brushed, yielding new groups which can be investigated using further features. Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
%
An hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection can also be derived from a shape-based clustering which creates shape groups.
%
Hypothesis-generation with focus on image-data starts with a shape based clustering.
%
The feature is mapped using a automatically chosen visualization appropriate for its data type.
%
The Visualization techniques have to combine both image-and non image data in order to set domain and range data in relation to each other.
%
Subsequently the visualization can be either brushed or new variables can be added to the analysis.
%
Brushes act as feature as they divide the subject space just like categorical variables and lead therefore to a new variable selection.
%
Selecting features also triggers a contingency analysis highlighting variables interacting with selected features, which is described in the following section.

@@

\subsection{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
%-------------------------------- Figure --------------------------------
% TODO: VISUALIZATION Figure sollte umgebaut werden, sodass sie die Schritte bei der Analyse hervorhebt
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/visualization}
 \caption{\emph{Image is not final} Screenshot from the front-end which is divided as follows: (a) Sidebar which contains all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table which shows exact numbers for each displayed variable combination. The data displayed is used to analyze the lumbar spine.
 }
 \label{fig:visualization}
\end{figure}
%-------------------------------- /Figure -------------------------------
%We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
%
%\emph{IVA} levels define different levels interaction. 
%
Suitability of an interaction- and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new interesting interactions.
%
%Since epidemiologists come from different expertise, their different background has to be taken into account.
%
The chances of a method to be adapted by epidemiologists increases highly if their domain language is used and the methods reflect routines they are used to.
%
Visual evaluation of data is therefore as important as methods allowing for numerical data analysis.
%
In the following sections we present the different parts of our proposed \emph{IVA} system for image-based cohort study analysis.
%Chosing or creating well suited visualization- and interaction techniques for epidemiological data
%The employed techniques need to respect the epidemiological requirements.

@@

\subsubsection{Structure of the system} \label{Structure and Workflow}
We divide the workspace into three major parts as seen in Figure~\ref{fig:visualization}.
\begin{itemize}
	\item The sidebar, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
	\item The canvas which holds all visualizations. Elements can be added, arranged, resized and removed freely.
	\item The interactive Pivot Table gives detailed numerical information of the features in the canvas view. This view on the data come natural to epidemiologists.
	\item The contingency view which depicts relations for features in the canvas.
\end{itemize}
%
The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.

Features can be dragged from the sidebar into the canvas area.
%
Doing so will automatically create an information visualization suitable for the current data type.
%
3D shape information about the investigated structure is displayed for each variable manifestation.
%
By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
%
Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
%
Shape based clustering can either be applied to all subjects or subgroups.

All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
%
Details to the different views are presented in the following sub-sections.

ToDo\\
- wie werden Confounder gefunden?\\
- welche statistischen Kennzahlen werden eingebunden?\\

\paragraph{Sidebar}
An overview of all variables is presented in a sidebar where they are categorized to different types such as somatometric, disease- or lifestyle related, pain indicators and laboratory data.
%
It also contains subject groups either defined by user brushing or by automated clustering.
%
Groups are treated exactly like other variables since they work exactly the same way which is dividing the subject space into different labeled categories.
% TODO Implement
Bar charts show the distribution of manifestations of each variable in the sidebar.

\paragraph{Adaptive Feature Visualization} \label{sec:AdaptiveFeatureVisualization}
Inspired be the previously discussed \emph{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
% TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
%
The 3D-plots can be navigated using standard mouse inputs and the camera is synchronized so that a direct comparison is given.
%
If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable. 
%
Each plot can be brushed using widgets.
%
It is able to duplicate brushes in order to create new groups which are evenly spread out.
%
A use case for this is when a continuous feature has to be divided into even groups.
\\
ToDo\\
- Dichotomous data\\
- Time-Line\\
- Statistical Analysis (Odds ratios)\\

TODO 2\\
- 3D Plots are small multiples - reference Tufte!

\paragraph{Pivot Tables}
Pivot tables are frequently used to present the data in epidemiological paper.
%When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
%
As seen in Figure~\ref{fig:visualization} (c) it is a good way to display how many subjects are in each group.
%
Pivot table quickly get confusing and cluttered when they are divided into to many subgroups.
%
We tackled this problem by making the order and number of displayed variables adaptable.
%
This also applies on the designation of Row or Column Variables.
% TODO IMPLEMENT
The mean shape for each resulting sub-group is also displayed for each subject.

\subsubsection{Automated Feature Suggestion}
% https://de.wikipedia.org/wiki/Kontingenzkoeffizient#Cram.C3.A9rs_V
TODO REWRITE USING CRAMERS V

As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
%
Turkay and colleagues used the approach to calculate various key figures based in the distribution functions of each feature derived from the image data \cite{Turkay2013}.
%
Since the majority of our data are categorical features, we have to employ different solutions.
%
A solution for this problem are odds ratios, which are a standard statistical tool for stating relations between features.
%
Odds ratios can only be calculated for $2\times2$ contingency tables which usually represent the presence of a condition in a population divided by a characteristic (e.g. male/female subjects with presence or absence of back pain).
%
To calculate odds ratios for variables with more than two manifestations, we calculate local odds ratios for each possible combination, yielding a matrix for each feature combination \cite{Rudas1998}.
%
When the subjects are divided into groups and the calculation is carried out again for all feature combination, the difference in sum of the odds weighted with the number of manifestations per feature can be used to indicate if the feature combination yields a difference.
%
These difference are then highlighted in a separate tab of the side bar "Interactions".
%
Interesting interactions then can be assessed creating a linked view using the standard drag and drop workflow.
\\\\
In the following sections we will discuss details on the implementation which relies on modern Web-Technologies.
\\\\
ToDo\\
- This can be improved--summing up the values possibly not the cleverest solution--calculation of variance etc. possible\\
- Matrix Visualization?

\subsection{Implementation}
\begin{figure}[htb]
 \centering
 \label{fig:technologies}
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/technologies}
 \caption{\emph{Image is not final}. The front-end solution (left) uses state of the art web technologies such as HTML5/CSS3, WebGL and SVG to display the data.
 % 
 The NodeJS based back-end (right) stores all image- and non-image data and transfers it to connected clients.
 %
 All computation heavy operations like calculation of mean-shapes or -distances as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
 %
 Client-Server communication is accomplished via the WebSocket protocol.
 }
\end{figure}
%
In order to provide a fast communication loop between method development and expert input, we decided to base all implementations on modern web technologies which benefits from various advantages:
\begin{itemize}
	\item No additional software needs to be installed, most people use decent state-of-the-art web browser, even on mobile devices.
	\item The Client-Server structure allows it to employ heavy computation on a server machine and transfer results to the client
	\item Since image data for several thousand subjects claims hundreds of Gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
	%
	High confidentiality standards of the data can be met by restricting access via a account system
	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web which results in many open source libraries which are well documented, rich in examples and driven by active communities.
\end{itemize}
These advantages do not come without drawbacks.
%
Many methods which specialized libraries like the Visualization Toolkit (VTK) or R for statistics have build in need to be written from scratch in order to fit in the context.
%

The back-end is written using Node, which based on the Google V8 Javascript runtime environment.
%
Due to its event-driven non-blocking I/O model it is fast and does not freeze in case of heavy workload like mesh calculation.
\\\\
Non-image data for all subjects including the data dictionary is stored on the Server in a JSON file.
%
Image data is available as raw DICOM files as well as segmented Meshes which can be used to compare subjects.
%
On client connection the requested files are transferred.
%
The server processes calculation heavy statistical tasks such as calculation of Odds Ratios or Chi Square tests for all variable combinations in order to keep the computation time on the client as low as possible.
%

The front-end is created using Twitter Bootstrap as foundation for the layout and basic UI elements using HTML5, CSS3 and Javascript.
%
Information Visualizations such as Scatterplots and bar charts are created using the popular Data Driven Documents library which works well for attaching data to visible elements like vector graphics.
%
WebGL rendering is done using the Threejs which allows GPU Accelerated data rendering.
%
Communication between Client and Server runs through the WebSockets protocol.
%
Since our clustering algorithms are written in MatLab we hat to access them using the Node Server.
%
We accomplished this by converting them to parameterized standalone console applications which are spawned by node on client request and then reads the result from the console standard-out and returns it in a proper format to the client.
%
All parameter steered console applications can be incorporated in this context.

@@

\paragraph{Image-Data.} \label{Image-Data}
The lumbar spine was detected in the image data using a hierarchical finite element method according to Rak and colleagues \cite{Rak2013}.
%
This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L3-vertebra.
%
Two user defined landmarks on the top and bottom of the L3-vertebra are used to obtain an initial height estimation of the model.
%Two more user- defined landmarks are the top and bottom of the L3-vertebra, which transforms size of the model accordingly to match the L3-vertebra .
%
It uses a weighted sum of T1- and T2-weighted MRI images to detect the lumbar spine shape.
%
The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
%
Due to TODO ANTWORT VON MARKO HIER, 983 models are obtained.
%
For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoliosis which are medical terms for spine curvature \cite{Klemm2013VMV}.

@@

Cluster 9 represents the subjects with average shape.
%
Other shapes differ with respect to size, such as Cluster 2, 8, 10 and 3 where the last one also is more straight, which is usual for subjects with larger body height.
%
Cluster 4, 7 and 5 contain outliers, characterized by their unusual shape and small number.
%
Cluster 8 was of special interest because of its large distance to the mean shape while still exposing the second highest subject count.
%
Looking at the \emph{Cram\'{e}rs' V} contingency values of the group reveals interactions of this group with employment status, body size, age, thyroid nodules and blood-fat value.
%
TODO Feedback einholen von Frau Hegenscheid, TODO Implement Color coding only in a certain direction!\\
TODO Selection of the variables and display them using the pivot table!
\\\\
While this approach does not assume any hypothesis, the data is met with a selection bias when compiling the list of related features by the domain experts.
%
It is arguable if this first filtering step, which is purely based on expert experience, is really a disadvantage, since it rules out many parameters which may interact with the presented features but the value of this knowledge would be small \cite{Wiley2008}.

>>>>

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
On the other hand, observations made by clinicians in the daily routine are translated into hypothesis for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical attributes which may influence a condition of interest.
%
The data variables necessary are gathered using structured interviews and clinical examinations.
%
Statistical methods like regression analysis aim to check the attribute list for plausibility.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Modern cohort studies often include medical image data.
%
These data needs to be segmented to analyze its features for correlations with non-image data.
%
Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques are more promising but challenging since the used methods as MRI and ultrasound are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.

@@

@@

@@

Morphometric variables are derived to allow for statistical comparison of the tissue which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed Version of the \emph{IVA Loop}. Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types. The data is visualized in the range and domain space, which can then be brushed, yielding new groups which can be investigated using further features. Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
%
An hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection can also be derived from a shape-based clustering which creates shape groups.
%
Hypothesis-generation with focus on image-data starts with a shape based clustering.
%
The feature is mapped using a automatically chosen visualization appropriate for its data type.
%
The visualization techniques have to combine both image-and non image data in order to set domain and range data in relation to each other.
%
Subsequently the visualization can be either brushed or new variables can be added to the analysis.
%
Brushes act as feature as they divide the subject space just like categorical variables and lead therefore to a new variable selection.
%
Selecting features also triggers a contingency analysis highlighting variables interacting with selected features, which is described in the following section.

@@

\subsection{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
%-------------------------------- Figure --------------------------------
% TODO: VISUALIZATION Figure sollte umgebaut werden, sodass sie die Schritte bei der Analyse hervorhebt
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/visualization}
 % \includegraphics[width=1\textwidth]{figures/visualization}
 \caption{\emph{Image is not final} Screenshot from the front-end which is divided as follows: (a) Sidebar which contains all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table which shows exact numbers for each displayed variable combination. The data displayed is used to analyze the lumbar spine.
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
%We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
%
%\emph{IVA} levels define different levels interaction. 
%
Suitability of an interaction- and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new interesting interactions.
%
%Since epidemiologists come from different expertise, their different background has to be taken into account.
%
The chances of a method to be adapted by epidemiologists increases highly if their domain language is used and the methods reflect routines they are used to.
%
Visual evaluation of data is therefore as important as methods allowing for numerical data analysis.
%
In the following sections we present the different parts of our proposed \emph{IVA} system for image-based cohort study analysis.
%Chosing or creating well suited visualization- and interaction techniques for epidemiological data
%The employed techniques need to respect the epidemiological requirements.

@@

\subsubsection{System Structure} \label{Structure and Workflow}
We divide the workspace into four major parts as seen in Figure~\ref{fig:visualization}.
\begin{itemize}
	\item The sidebar, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
	\item The canvas which holds all visualizations. Elements can be added, arranged, resized and removed freely.
	\item The interactive Pivot Table gives detailed numerical information of the features in the canvas view. This view on the data come natural to epidemiologists.
	\item The contingency view depicting relations for features in the canvas.
\end{itemize}
%
% TODO Include:
% The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
% 
% Features can be dragged from the sidebar into the canvas area.
% %
% Doing so will automatically create an information visualization suitable for the current data type.
% %
% 3D shape information about the investigated structure is displayed for each variable manifestation.
% %
% By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
% %
% Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
% %
% Shape based clustering can either be applied to all subjects or subgroups.
% 
% All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
% %
% Details to the different views are presented in the following sub-sections.

\paragraph{Sidebar.}
An overview of all features is presented in a sidebar where they are categorized to different types such as somatometric, disease- or lifestyle related, pain indicators and laboratory data.
%
It also contains subject groups either defined by brushing or automated shape-clustering.
%
Groups are treated like features since they work exactly the same way by dividing the subject space into labeled categories.
%
Features can be dragged from the sidebar into the canvas area.
%
This triggers an adaptive feature visualization suitable for the current data type.
%Bar charts show the distribution of manifestations of each variable in the sidebar.

\paragraph{Adaptive Feature Visualization.} \label{sec:AdaptiveFeatureVisualization}
Inspired be previously the discussed \texttt{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
%
% TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
Following Tuftes concept of \emph{small multiples} \cite{Tufte1983} the medical image data is directly included into the plot by including color coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
%If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
%
The 3D-plots can be navigated using standard mouse inputs, the camera is synchronized between all views to enable direct comparison.
%
Mapping the distance from a group mean shape to the global mean to color allows to assess local shape changes.
%%TODO Adaptive visualization!
If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable. 
%
Each plot can be brushed using widgets.
%
Brush selections are propagated to all visualizations allowing for fast feature querying.
% TODO Timeline
%
%It is possible to duplicate brushes in order to create new groups which are evenly spread out.
%
%A use case for this is when a continuous feature has to be divided into even groups.

\paragraph{Pivot Tables.}
Pivot tables are frequently used to present the data in epidemiological publications.
%
Epidemiologists are used to process groups based on table representations so decided to introduce an interactive pivot table.
%When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
%
As seen in Figure~\ref{fig:visualization} (c) it is a good way to display how many subjects are in each group.
%
Pivot table quickly get confusing and cluttered when they are divided into to many subgroups.
%
We tackled this problem by making the order and number of displayed variables adaptable.
%
This also applies on the designation of Row or Column Variables.
%
Another way to avoid clutter is the user-driven selection of displayed variables.
%
To allow better comparison with respect to features the values of each cell can also be displayed as percentage of the feature represented of either the row or the column.
% TODO IMPLEMENT
%The mean shape for each resulting sub-group is also displayed for each subject.

\paragraph{Automated Feature Suggestion using a Contingency Matrix.}
As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
%
Turkay and colleagues used the approach to calculate various key figures based in the distribution functions of each feature derived from the image data \cite{Turkay2013}.
%
Since the majority of our data are categorical features, we have to employ different solutions.
%
The \emph{Cram\'{e}rs' V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
%
It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two variables.
%
\emph{Cram\'{e}rs' V} is defined as:
\begin{equation}
V = \sqrt{\frac{X^2}{N(k-1)}},
\end{equation}
where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total of observations and $k$ either the row or column count, depending on which one is lower.
%
$V$ assumes values between $0$, meaning two variables are completely independent and $1$ indicating they are the same.
%
\emph{Cram\'{e}rs' V} is always positive and does therefore not allow statements about dependency direction.

It shares the same restrictions as \emph{Pearson's $X^2$} which include next to others expected count larger than 5 for $80\%$ of the contingency table entry and no expected value can be smaller than one.
%
Therefore are manifestations who are only exposed by small subject groups not assessable using this technique.
%
This makes them not assessable for epidemiological analysis as well since statistical validation needs a minimum count to be seen as valid.
% Limitations and usual space
The contingency matrix highlights contingencies of visualized features with features in the data set.
%
This aims to highlight features which may interact with the focused hypothesis as well as trigger new hypotheses.
%
Contingency is visualized using an interactive adjacency matrix with association power mapped to color hue.
%
The distinction whether a interaction is a confounder or and effect depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
\\\\
In the following Section we discuss how we implemented the presented methods using free on open Web standards.
% \subsubsection{Automated Feature Suggestion}
% % https://de.wikipedia.org/wiki/Kontingenzkoeffizient#Cram.C3.A9rs_V
% 
% As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
% %
% Turkay and colleagues used the approach to calculate various key figures based in the distribution functions of each feature derived from the image data \cite{Turkay2013}.
% %
% Since the majority of our data are categorical features, we have to employ different solutions.
% %
% A solution for this problem are odds ratios, which are a standard statistical tool for stating relations between features.
% %
% Odds ratios can only be calculated for $2\times2$ contingency tables which usually represent the presence of a condition in a population divided by a characteristic (e.g. male/female subjects with presence or absence of back pain).
% %
% To calculate odds ratios for variables with more than two manifestations, we calculate local odds ratios for each possible combination, yielding a matrix for each feature combination \cite{Rudas1998}.
% %
% When the subjects are divided into groups and the calculation is carried out again for all feature combination, the difference in sum of the odds weighted with the number of manifestations per feature can be used to indicate if the feature combination yields a difference.
% %
% These difference are then highlighted in a separate tab of the side bar "Interactions".
% %
% Interesting interactions then can be assessed creating a linked view using the standard drag and drop workflow.
% \\\\
% In the following sections we will discuss details on the implementation which relies on modern Web-Technologies.
% \\\\
% ToDo\\
% - This can be improved--summing up the values possibly not the cleverest solution--calculation of variance etc. possible\\
% - Matrix Visualization?

\subsection{Implementation}
\begin{figure}[htb]
 \centering
 \label{fig:technologies}
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/technologies}
 \caption{\emph{Image is not final}. The front-end solution (left) uses state of the art web technologies such as HTML5/CSS3, WebGL and SVG to display the data.
 % 
 The NodeJS based back-end (right) stores all image- and non-image data and transfers it to connected clients.
 %
 All computation heavy operations like calculation of mean-shapes or -distances as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
 %
 Client-Server communication is accomplished via the WebSocket protocol.
 }
\end{figure}
%
To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies which benefits from various advantages:
\begin{itemize}
	\item No additional software needs to be installed, most people use decent state-of-the-art web browser, even on mobile devices.
	\item The client-server structure allows for employing heavy computation on a server machine and transfer results to the client.
	\item Since image data for several thousand subjects claims hundreds of gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
	%
	High confidentiality standards of the data can be met by restricting access via a account system
	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web which results in many open source libraries which are well documented, rich in examples and driven by active communities.
	\item Web technologies are free.
\end{itemize}
These advantages do not come without drawbacks.
%
Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
%

The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \url{nodejs.org}}, which based on the Google V8 Javascript runtime environment.
%
Due to its event-driven non-blocking I/O model it is fast and does not freeze in case of heavy workload like mesh calculation.
\\\\
Non-image data for all subjects including the data dictionary is stored in a JSON file on the Server.
%
Image data is available as raw DICOM files as well as segmented meshes, which can be used to compare subjects.
%
On client connection the requested files are transferred.
%
The server processes calculation heavy statistical tasks such as calculation of \emph{Cram\'{e}rs' V} values for all variable combinations in order to keep the computation time on the client as low as possible.
%

The front-end is created using \texttt{Twitter Bootstrap}\footnote{Developed by Twitter, \url{getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%
Information Visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} \cite{D3} library, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
%
\texttt{WebGL} rendering is done using the \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{threejs.org}} and allows GPU Accelerated data rendering.
%
Communication between Client and Server runs through the WebSockets protocol.
%
Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{www.mathworks.com}} we hat to access them using the \texttt{NodeJS} Server.
%
We accomplished this by converting it to a parameterized standalone console application, that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard-out and returns it in a proper format to the client.
%
All parameter steered console applications can be incorporated in this context.

The following section describes how our presented \emph{IVA} workflow works for a epidemiological use case.

@@

\paragraph{Image-Data.} \label{Image-Data}
The lumbar spine was detected in the image data using a hierarchical finite element method according to Rak and colleagues \cite{Rak2013}.
%
This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L3-vertebra.
%
Two user defined landmarks on the top and bottom of the L3-vertebra are used to obtain an initial height estimation of the model.
%Two more user- defined landmarks are the top and bottom of the L3-vertebra, which transforms size of the model accordingly to match the L3-vertebra .
%
It uses a weighted sum of T1- and T2-weighted MRI images to detect the lumbar spine shape.
%
The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
%
Due to incorrect initialization, strongly deformed spines, contrast-differences and artifacts, the model was not able to detect lumbar spines for all subjects.
%
We obtained and work with 983 models.
%
For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoliosis which are medical terms for spine curvature \cite{Klemm2013VMV}.

@@

Cluster 9 represents the subjects with average shape.
%
Other shapes differ with respect to size, such as Cluster 2, 8, 10 and 3 where the last one also is more straight, which is usual for subjects with larger body height.
%
Cluster 4, 7 and 5 contain outliers, characterized by their unusual shape and small number.
%
Cluster 8 was of special interest because of its large distance to the mean shape while still exposing the second highest subject count.
%
Looking at the \emph{Cram\'{e}rs' V} contingency values of the group reveals interactions of this group with employment status, body size, age, thyroid nodules and blood-fat value.
%
TODO Feedback Greifswald, 
%TODO Implement Color coding only in a certain direction!\\
%TODO Selection of the variables and display them using the pivot table!
\\\\
While this approach does not assume any hypothesis, the data is met with a selection bias when compiling the list of related features by the domain experts.
%
It is arguable if this first filtering step, which is purely based on expert experience, is really a disadvantage, since it rules out many parameters which may interact with the presented features but the value of this knowledge would be small \cite{Wiley2008}.
